---
title: "Differential Abundance in the ocular microbiome in DED"
author: "Mary Bearkland"
date: "02DEC2024"
output:
  pdf_document: default
  html_document: default
  
    
---

```{r}
# Set a CRAN mirror
options(repos = c(CRAN = "https://cran.rstudio.com"))

install.packages("xfun")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(results = 'hide', fig.show = 'hide')
```

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(c("devtools", "tidyverse", "kableExtra"))
BiocManager::install(c("waldronlab/bugSigSimple", "waldronlab/BugSigDBStats", "waldronlab/bugsigdbr"))
```

```{r}
suppressPackageStartupMessages({
  library(bugSigSimple)
  library(BugSigDBStats)
  library(bugsigdbr)
  library(tidyverse)
  library(stringr)
  library(kableExtra)
  library(dplyr)
  library(magrittr)
  library(boot)
})
```

use version="devel" and cache = FALSE to take the latest version from bugsigdb.org

```{r}
dat <- bugsigdbr::importBugSigDB(version = "devel", cache = FALSE) 
dim(dat)
```

Examine the data pulled in to view variable names

```{r}
names(dat)
```

Subset to include only PMID numbers of interest

```{r}
included.pmid <-
  c(
    31463790,
    32931939,
    37586456, 
    37762390,
    34900990,
    37803284,
    35350577,
    37026303,
    32694705,
    38111925
  )
subset.dat <-
  dat[dat$PMID %in% included.pmid,] 
```

```{r}
dim(subset.dat)
```

Confirm all studies included

```{r}
subset.dat
```

Examine group 0 names

```{r}
unique(subset.dat$`Group 0 name`)
```

Examine group 1 names

```{r}
unique(subset.dat$`Group 1 name`)
```

To omit the double comparison in Study 856 by removing the ANOSIM analysis

```{r}
subset.dat <- subset.dat %>%
  filter(`Statistical test` != "ANOSIM")
subset.dat
```

Final subset to include only those Group 0 names of interest (controls with no DED) and only those Group 1 names of interest (some variation on DED)

```{r}
subset.final <- subset.dat[which(subset.dat$`Group 0 name` %in% c("Healthy Control", "Control", "control", "healthy controls", "Normal Control (NC)","Healthy control", "Healthy Controls","Normal healthy (NDM) children", "Healthy controls")& subset.dat$`Group 1 name`%in% c("Meibomian Gland Dysfunction + Lacrimal Dysfunction", "Meibomian Gland Dysfunction", "Dry Eye","ADDE", "DED patients", "MGD", "Meibomian Gland Dysfunction (MGD)", "Meibomian Gland Dysfunction (MGD) DED", "Meibomian Gland Dysfunction (MGD) Groups", "Mixed DED", "Sjogrens Syndrome Dry Eye (SSDE)", "Non Sjogrens Syndrome Dry Eye (NSSDE)", "Dry Eye Disease patients", "Sjogren's patients with Dry Eye Disease", "Dry Eye Disease patients without Sjogrens", "Diabetic children with Dry Eye Disease (DM-DE)", "Mild Dry Eye", "Mild and Moderate to Severe Dry Eye")),]
subset.final
```

```{r}
dim(subset.final)
```

These are the studies included in the review

```{r}
bugSigSimple::createStudyTable(subset.final)|> kableExtra::kbl()
```

This table summarizes the results for the identified taxa.

```{r}
# Install and load necessary packages
install.packages("openxlsx")
library(openxlsx)
library(knitr)

# Create and format the table
taxon_table <- bugSigSimple::createTaxonTable(subset.final, n = 2000)
styled_table <- kable_styling(kbl(taxon_table))

# Create a new workbook
wb <- createWorkbook()

# Add a worksheet
addWorksheet(wb, "Taxon Table")

# Write the data to the worksheet
writeData(wb, "Taxon Table", taxon_table)

# Save the workbook
saveWorkbook(wb, "TaxonTable.xlsx", overwrite = TRUE)

```

```{r}
install.packages("writexl")

```

```{r}
install.packages("ontologyIndex")
```

```{r}
efo<-getOntology("efo")
efo
```

Top 1000 taxa decreased in relation to the control

```{r}
mostfreqdec<-bugSigSimple::getMostFrequentTaxa(subset.final, n=1000,sig.type = "decreased")
```

```{r}
mostfreqdec
```

Top 1000 taxa increased in relation to the control

```{r}
mostfreqinc<-bugSigSimple::getMostFrequentTaxa(subset.final,n=1000, sig.type = "increased")
mostfreqinc
```

To obtain signatures

```{r}
allsigs <- bugsigdbr::getSignatures(subset.final, tax.id.type = "taxname")
allsigs <- allsigs[sapply(allsigs, length) > 0] #require length > 0
dim(allsigs)
```

To obtain signature lengths displayed in a bar graph and import to a word document

```{r}
# Install and load necessary packages

library(ggplot2)


# Create the bar graph
# Calculate the lengths of signatures
siglengths <- sapply(allsigs, length)

# Create a dataframe
siglengths.df <- data.frame(siglengths = siglengths)

# Plot the bar graph with Y-axis label
plot <- ggplot(siglengths.df, aes(x = siglengths)) +
  geom_bar() +
  labs(y = "Number of signatures", x = "Signature Lengths")

# Display the plot
print(plot)


# Save the plot to a file
ggsave("bar_graph.png", plot)


```

Table of signature lengths

```{r}
table(siglengths)
```

Calculate pairwise overlap and obtain Jaccard value to determine similarity of signatures

```{r}
mydists <- BugSigDBStats::calcPairwiseOverlaps(allsigs)
dim(mydists)
```

```{r}
library(grid)
library(ComplexHeatmap)

jmat <- BugSigDBStats::calcJaccardSimilarity(allsigs)

# Truncate row names at the first underscore
rownames(jmat) <- sub("_.+$", "", rownames(jmat))

# Truncate column names at the first underscore
colnames(jmat) <- sub("_.+$", "", colnames(jmat))

# Check the truncated names
cat("Truncated row names: ", rownames(jmat), "\n")
cat("Truncated column names: ", colnames(jmat), "\n")

# Ensure row_labels and column_labels lengths match the matrix dimensions
row_labels <- rownames(jmat)
col_labels <- colnames(jmat)

# Verify that lengths match
cat("Length of row_labels: ", length(row_labels), "\n")
cat("Length of col_labels: ", length(col_labels), "\n")

# Create custom annotations with line breaks
ha <- HeatmapAnnotation(
  `Signature\nLength` = anno_barplot(siglengths, gp = gpar(fill = 4:13), 
                                      annotation_name_gp = gpar(fontsize = 10, lineheight = 0.8))
)
hr <- rowAnnotation(
  `Signature\nLength` = anno_barplot(siglengths, gp = gpar(fill = 4:13), 
                                      annotation_name_gp = gpar(fontsize = 10, lineheight = 0.8))
)

# Adjust heatmap dimensions and settings
hm <- Heatmap(
  jmat,
  top_annotation = ha, 
  left_annotation = hr,
  row_names_max_width = unit(10, "cm"),  # Adjusted width
  column_names_max_height = unit(3, "cm"),  # Adjusted height for column names
  row_labels = row_labels,
  column_labels = col_labels,
  column_title = "Fig 3 Heatmap of Similarity of Signatures Between Studies",  # Title of the heatmap
  column_title_gp = gpar(fontsize = 15),  # Adjust font size for title
  heatmap_legend_param = list(
    title = "Legend", 
    direction = "horizontal"  # Set legend direction to horizontal
  )  # Customizing legend
)

# Draw the heatmap
draw(hm, 
     heatmap_legend_side = "bottom", 
     annotation_legend_side = "right")

# Save the heatmap as an image
png("heatmap.png", width = 12, height = 8, units = "in", res = 300)
draw(hm, 
     heatmap_legend_side = "bottom", 
     annotation_legend_side = "right")
dev.off()



To better visualize

```{r}
library(InteractiveComplexHeatmap)
hm <- draw(hm)
htShiny(hm)
```

```{r}
library(ComplexHeatmap)
library(grid)
library(officer)

# Assuming jmat and other variables are already defined as before

# Perform hierarchical clustering
hc <- hclust(dist(jmat))
plot(hc)

# Save the cluster map as an image
png("cluster_map.png", width = 12, height = 8, units = "in", res = 300)
plot(hc, main = "Hierarchical Cluster Map of Signatures", xlab = "", sub = "", cex = 0.6)
dev.off()


```

```{r}
clusts <- sort(cutree(hc, h = 0.05))
lapply(unique(clusts), function(i) names(clusts)[clusts == i])
```

```{r}
#all taxon increased and decreased freq
allfreqs <- bugSigSimple::createTaxonTable(subset.final, n = 500) %>% #could change number
  arrange(I(decreased_signatures - increased_signatures))
incfreqs <- filter(allfreqs, I(increased_signatures - decreased_signatures) > 0)
decfreqs <- filter(allfreqs, I(increased_signatures - decreased_signatures) < 0)
kableExtra::kbl(allfreqs) %>%
  kable_paper("hover", full_width = FALSE)
```
