---
title: "Delivery Mode and the Meconium Microbiome"
name: "Shaimaa Elsafoury"
date: "2/24/2022"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Delivery Mode and the Meconium Microbiome}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Analysis objectives

1. Import, recode, and subset data from bugsigdb.org    
2. Create a table of studies
3. Calculate the frequency of appearance of each taxa in independent signatures and identify the most frequently reported taxa

# Making sure packages are installed

Not evaluated in vignette:
```{r, eval=FALSE, messages=FALSE}
install.packages(c("devtools", "tidyverse", "kableExtra"))
devtools::install_github("waldronlab/bugSigSimple")
devtools::install_github("waldronlab/BugSigDBStats")
devtools::install_github("waldronlab/bugsigdbr")
```

# Load and subset data

```{r, messages=FALSE}
suppressPackageStartupMessages({
  library(bugSigSimple)
  library(BugSigDBStats)
  library(bugsigdbr)
  library(tidyverse)
  library(stringr)
  library(kableExtra)
  library(dplyr)
})
```

```{r}
# use version="devel" and cache = FALSE to take the latest version from bugsigdb.org
dat <- bugsigdbr::importBugSigDB(version = "devel", cache = FALSE) 
dim(dat)
names(dat)
```

# Subsetting

Subsetting only studies done on humans - 6 of 17 studies were excluded because they were either done on mice or not statistically reliable (Cregger et. al)

```{r, messages=FALSE}
included.pmid <-
  c(
    28018325,
    24614698,
    29207565,
    29459704,
    29538354,
    32012716,
    20566857,
    28512451,
    28112736,
    27362264
  )
subset.dat <-
  filter(dat, PMID %in% included.pmid) 
```

```{r group0}
unique(subset.dat$`Group 0 name`)
included.group0 <-
  c(
    "vaginal delivery",
    "vaginally born controls",
    "vaginal delivery in mecomium",
    "vaginally born controls"
  )
```

```{r group1}
unique(subset.dat$`Group 1 name`)
included.group1 <-
  c("c-section",
    "C-section",
    "c-section delivery",
    "C-section delivery")
```

```{r}
subset.final <-
  filter(subset.dat, `Group 0 name` %in% included.group0 & `Group 1 name` %in% included.group1) %>%
  filter(`Body site` == "meconium") %>%
  arrange(PMID)
select(subset.final, all_of(c("PMID", "Source", "Group 0 name", "Group 1 name", "Abundance in Group 1")))
```

# Summary of studies

```{r, messages=FALSE}
bugSigSimple::createStudyTable(subset.final)
```

# Most frequently identified taxa

```{r}
getMostFrequentTaxa(subset.final, n = 50)
```

```{r}
getMostFrequentTaxa(subset.final, sig.type = "increased")
```

```{r}
getMostFrequentTaxa(subset.final, sig.type = "decreased")
```
