---
title: "COVID-19 associated microbiome"
author: "Clare Grieve"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{COVID-19 associated microbiome}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Analysis objectives

1. Import, recode, and subset data from bugsigdb.org    
2. Create a table of studies
3. Create a clustered heatmap showing similarity of signatures from independent studies
4. Calculate the frequency of appearance of each taxa in independent signatures, and identify the most frequently reported taxa
5. Estimate the probability of the most frequently identified taxa occuring so frequently by chance

# Packages installation

Install packages (not evaluated in vignette)

```{r, eval=FALSE}
install.packages(c("devtools", "tidyverse", "kableExtra", "gt", "glue"))
devtools::install_github("waldronlab/bugSigSimple")
devtools::install_github("waldronlab/BugSigDBStats")
devtools::install_github("waldronlab/bugsigdbr")
```

# Data import, recoding, and subset

```{r}
library(bugSigSimple)
dat <- bugsigdbr::importBugSigDB(cache = TRUE) 
dim(dat)
names(dat)
```

```{r}
library(tidyverse)
 efo <- bugsigdbr::getOntology("efo")
covid_all <- bugsigdbr::subsetByOntology(dat, column = "Condition", "COVID-19", efo) %>%
  mutate(studyexp = paste(Study, Experiment, sep = "_")) %>%
  mutate(
    site = recode(`Body site`,
      "Feces" = "Gut",
      "Rectum" = "Gut",
      "Nasopharynx" = "aURT",
      "Oropharynx" = "aURT",
      "Nasopharynx,Oropharynx" = "aURT",
      "Nasal cavity" = "aURT",
      "Surface of tongue" = "aURT",
      "Nasopharynx,Throat" = "aURT",
      "Throat" = "aURT",
      "Tongue" = "aURT",
      "Lung" = "LRT",
      "Sputum" = "LRT"
    )
  ) %>%
  mutate(comparison1 = paste(`Group 0 name`, `Group 1 name`, sep = " vs "))
```  

## Table of studies

```{r}
bugSigSimple::createStudyTable(covid_all)
```


## Taxon frequency tables by body site

```{r}
library(dplyr)
gut_sigs <- filter(covid_all, 
                           site == "Gut") %>%
    drop_na(Source)

naso_sigs <- filter(covid_all, 
                           site == "aURT") %>%
    drop_na(Source)
  
resp_sigs <- filter(covid_all, 
                           site == "LRT") %>%
  drop_na(Source)
```

In this table, the Binomial Test p-value corresponds to the null hypothesis

H0: the proportion of signatures in which the taxon is reported increased or decreased, relative to the total number of signatures in which it is reported, is equal to 0.5

```{r}
kableExtra::kbl(bugSigSimple::createTaxonTable(gut_sigs))
```

```{r}
kableExtra::kbl(bugSigSimple::createTaxonTable(naso_sigs))
```

```{r}
kableExtra::kbl(bugSigSimple::createTaxonTable(resp_sigs))
```

# gut microbiota analysis

Look specifically at case-control comparisons
```{r}
healthy <- grepl(".*(healthy.*COVID|COVID.*healthy).*", gut_sigs$comparison1, ignore.case = TRUE)

cc_gut_sigs1 <- gut_sigs %>%
      filter(healthy == TRUE) 
```

## Overall frequencies of taxa increased in cases for case/control feces studies

Identifying a taxon reported consistently in 8 out of 8 studies is _much_ more compelling that the FDR value here would suggest, since this taxon also passed a significance threshold in every one of those studies.

```{r}
cc_gut_sigs1_taxontable <- bugSigSimple::createTaxonTable(cc_gut_sigs1) %>% 
  mutate(FDR =  p.adjust(p = `Binomial Test pval`, method="fdr")) %>%
  relocate(FDR, .after = `Binomial Test pval`)
kableExtra::kbl(cc_gut_sigs1_taxontable)
```

## Monte-Carlo simulation for increased abundance taxa

Just for the increased cc_gut_sigs1 for now. I am inclined to skip this analysis in favor of the clustering and binomial test analysis.

```{r}
#library(bugSigSimple)
#gut.sigs.increased <- filter(cc_gut_sigs1, `Abundance in Group 1` == "increased") %>% 
 # bugsigdbr::getSignatures(tax.id.type = "taxname")
# my.siglengths.inc <- sapply(gut.sigs.increased, length)

#getCriticalN(gut.sigs.increased, my.siglengths.inc)
# Compare to observed - enterococcus & streptococcus are the only taxa that equal the critical limit
#frequencySigs(gut.sigs.increased)
```

```{r}
library(bugSigSimple)
library(ggplot2)

gut.sigs.increased <- filter(cc_gut_sigs1, `Abundance in Group 1` == "increased") %>%
  bugsigdbr::getSignatures(tax.id.type = "taxname")

my.siglengths.inc <- sapply(gut.sigs.increased, length)
```


```{r}
cn_res <- getCriticalN(
  relevant.sigs = gut.sigs.increased,
  siglengths    = my.siglengths.inc,
  alpha         = 0.05,
  nsim          = 1000,   
  ci            = 0.95,
  ci_nsim       = 500
)

cat(sprintf(
  "Critical N (%s): %d\n95%% CI: [%d, %d]\n",
  cn_res$quantile_label,
  round(cn_res$critical_n),
  round(cn_res$ci_lower),
  round(cn_res$ci_upper)
))
```

```{r}

shapiro.test(cn_res$simulated_values)
obs_max <- max(table(unlist(gut.sigs.increased)))

plotCriticalN(
  cn_result = cn_res,
  obs_max   = obs_max,
  title     = "Normal distribution — max taxon frequency\n(gut microbiome, COVID-19 increased taxa)"
)
```
```{r significant-taxa-plot, fig.width=11, fig.height=nrow(sig_df) * 0.55 + 3}
sig_df <- significantTaxa(
  sigs      = gut.sigs.increased,
  cn_result = cn_res,
  n         = 20,
  plot      = FALSE,      # suppress internal plot so we can correct significance first
  plot_type = "lollipop",
  title     = "Taxon frequency vs. Monte Carlo threshold\n(gut, COVID-19 increased)"
)
```


```{r significant-taxa-plot, fig.width=11, fig.height=nrow(sig_df) * 0.55 + 3}
cat("Critical N value:  ", cn_res$critical_n, "\n")
cat("Max observed count:", max(sig_df$count), "\n")
print(sig_df[, c("taxon", "count", "critical_n")])
```

```{r significant-taxa-plot, fig.width=11, fig.height=nrow(sig_df) * 0.55 + 3}
# Correct significance flag from > to >= to include taxa that meet the threshold
sig_df$significant <- sig_df$count >= sig_df$critical_n
```

```{r}
cat("\nSignificant taxa:\n")
print(sig_df[sig_df$significant, c("taxon", "count", "critical_n", "significant")])
```

```{r significant-taxa-plot, fig.width=11, fig.height=nrow(sig_df) * 0.55 + 3}
# Re-order factor for clean top-to-bottom ranking
sig_df$taxon <- factor(sig_df$taxon, levels = sig_df$taxon[order(sig_df$count)])
```


```{r significant-taxa-plot, fig.width=11, fig.height=nrow(sig_df) * 0.55 + 3}
# Now render the lollipop plot with corrected significance
ggplot(sig_df, aes(x = count, y = taxon, colour = factor(significant))) +

  annotate("rect",
           xmin = cn_res$ci_lower, xmax = cn_res$ci_upper,
           ymin = -Inf, ymax = Inf,
           fill = "grey70", alpha = 0.15) +

  geom_vline(xintercept = cn_res$critical_n,
             linetype = "dashed", colour = "grey30", linewidth = 0.8) +

  geom_segment(aes(x = 0, xend = count, yend = taxon),
               linewidth = 0.7, alpha = 0.5) +

  geom_point(size = 5) +

  geom_text(data = subset(sig_df, significant),
            aes(label = paste0("p = ", signif(p_empirical, 2))),
            hjust = -0.25, size = 3.5, colour = "#c0392b", fontface = "bold") +

  annotate("text",
           x = cn_res$critical_n, y = 0.4,
           label = paste0("Critical N = ", round(cn_res$critical_n)),
           hjust = 0.5, vjust = 1, size = 3.2, colour = "grey30") +

  scale_colour_manual(
    values = c("FALSE" = "#2980b9", "TRUE" = "#c0392b"),
    labels = c("FALSE" = "Not significant",
               "TRUE"  = paste0("Significant (>= Critical N = ",
                                round(cn_res$critical_n), ")")),
    name = NULL
  ) +

  scale_x_continuous(expand = expansion(mult = c(0, 0.25))) +
  scale_y_discrete(expand = expansion(add = c(1.2, 1.2))) +

  labs(
    title    = "Taxon frequency vs. Monte Carlo threshold\n(gut, COVID-19 increased)",
    subtitle = paste0("Dashed line = critical N (", cn_res$quantile_label,
                      ");  grey band = 95% CI"),
    x        = "Observed frequency (number of signatures)",
    y        = NULL
  ) +

theme_bw(base_size = 13) +
  theme(
    legend.position    = "bottom",
    plot.title         = element_text(face = "bold"),
    plot.subtitle      = element_text(colour = "grey40"),
    axis.text.y        = element_text(
      size   = 11,
      face   = ifelse(levels(sig_df$taxon) %in%
                        sig_df$taxon[sig_df$significant],
                      "bold.italic", "italic"),
      margin = margin(r = 4)
    ),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    plot.margin        = margin(t = 15, r = 30, b = 15, l = 15)
  )
```


```{r significant-taxa-table}
# Print the significance table (significant taxa first)
kableExtra::kbl(
  sig_df,
  col.names = c("Taxon", "Observed count", "Critical N",
                "Significant", "Empirical p-value"),
  digits    = 3,
  caption   = "Top 20 taxa: observed frequency vs. Monte Carlo threshold"
) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover")) %>%
  kableExtra::row_spec(
    which(sig_df$significant),
    bold       = TRUE,
    background = "#fde8e8"
  )
```

## Overall frequencies of taxa decreased in cases for case/control feces studies
```{r, eval = FALSE}
createTaxonTable(cc_gut_sigs1, n=40)
```

## nasopharyngeal microbiota analysis

Look specifically at case-control comparisons

```{r}
library(dplyr)
healthy <- grepl(".*(control.*COVID|COVID.*control).*", naso_sigs$comparison1, ignore.case = TRUE)

cc_naso_sigs1 <- naso_sigs %>%
      filter(healthy == TRUE) %>%
  subset(Study != "Study 458")
```

# Cluster analysis

Note, this EDA should really be done _before_ hypothesis testing. 

First calculate pairwise overlaps for all signatures of length > 1:
```{r}
allsigs <- bugsigdbr::getSignatures(covid_all, tax.id.type = "taxname")
allsigs <- allsigs[sapply(allsigs, length) > 1] #require length > 1
dim(allsigs)
mydists <- BugSigDBStats::calcPairwiseOverlaps(allsigs)
dim(mydists)
```

Create a matrix of Jaccard similarities (0 for no overlap, 1 for 100% overlap)

```{r}
jmat <- BugSigDBStats::calcJaccardSimilarity(allsigs)
```

```{r, fig.width=20, fig.height=20}
library(ComplexHeatmap)
siglengths <- sapply(allsigs, length)
ha <- HeatmapAnnotation(`Signature Length` = anno_barplot(siglengths))
hr <- rowAnnotation(`Signature Length` = anno_barplot(siglengths))
hm <- Heatmap(
  jmat,
  top_annotation = ha, left_annotation = hr,
  row_names_max_width = unit(20, "cm"),
  column_names_max_height = unit(20, "cm"),
  row_labels = sub(".+:", "", rownames(jmat)),
  column_labels = sub(".+:", "", colnames(jmat))
)
hm
```

Use this interactively to make an interactive heatmap (not evaluated in vignette). Some expanding of the default size is required to see anything. Creating a sub-heatmap, then exporting it as a table, allows in-depth identification of the subgroups.

```{r, eval = FALSE}
#library(InteractiveComplexHeatmap)
#hm <- draw(hm)
#htShiny(hm)
```

```{r}
library(InteractiveComplexHeatmap)

ht_jaccard <- draw(
  hm,
  heatmap_legend_side = "bottom",
  annotation_legend_side = "bottom"
)

htShiny(
  ht_jaccard,
  width1  = 1000,   # main heatmap panel width  (px)
  height1 = 1000,   # main heatmap panel height (px)
  width2  = 400,    # sub-heatmap panel width   (px)
  height2 = 400     # sub-heatmap panel height  (px)
)
```


```{r, fig.height=60, fig.width=20}
hc <- hclust(as.dist(jmat))
plot(hc)
```

This tree can be cut to show the clusters, for example. The clusters of more than 1 signature but less than ~10 signatures are most likely to be something interesting.

```{r}
clusts <- sort(cutree(hc, h = 0.05))
lapply(unique(clusts), function(i) names(clusts)[clusts == i])
```

# Create a wide-format dataframe

This would be suitable for regression analysis.

```{r}
covid_withsigs <- filter(covid_all, !is.na(covid_all$`NCBI Taxonomy IDs`))
sigs <- bugsigdbr::getSignatures(covid_withsigs, tax.id.type = "taxname")
cmat <- t(safe::getCmatrix(sigs, as.matrix = TRUE, min.size = 0, prune = FALSE))
cdf <- data.frame(cmat, stringsAsFactors = FALSE, check.names = FALSE)
cdf <- cbind(covid_withsigs, cdf)
colnames(cdf)[1:54]
```

Note this has a number of columns that are mostly zeros, it could be filtered significantly for any regression or machine learning analysis:

```{r}
table(cdf[["[Brevibacterium] frigoritolerans"]])
```

Create another heatmap on correlations of presence/absence of taxa. This is not necessary because the previous Jaccard Index heatmap is probably better, it is just a demonstration of doing something with the taxa presence/absence directly. 

```{r, fig.width=20, fig.height=20}
sigcors <- cor(t(cmat))
siglengths <- sapply(sigs, length)
ha <- HeatmapAnnotation(`Signature Length` = anno_barplot(siglengths))
hr <- rowAnnotation(`Signature Length` = anno_barplot(siglengths))
hm <- Heatmap(
  sigcors,
  top_annotation = ha, left_annotation = hr,
  row_names_max_width = unit(20, "cm"),
  column_names_max_height = unit(20, "cm"),
  row_labels = sub(".+:", "", rownames(sigcors)), ##removing study just to make signature names legible
  column_labels = sub(".+:", "", colnames(sigcors))
)
hm
```

Use this interactively to make an interactive heatmap:
```{r, eval = FALSE}
#library(InteractiveComplexHeatmap)
#hm <- draw(hm)
#htShiny(hm)
```

```{r}
library(InteractiveComplexHeatmap)

ht_jaccard <- draw(
  hm,
  heatmap_legend_side = "bottom",
  annotation_legend_side = "bottom"
)

htShiny(
  ht_jaccard,
  width1  = 1000,   # main heatmap panel width  (px)
  height1 = 1000,   # main heatmap panel height (px)
  width2  = 400,    # sub-heatmap panel width   (px)
  height2 = 400     # sub-heatmap panel height  (px)
)
```


