---
title: "feces sigcomp"
author: "clare"
date: "8/2/2021"
output: html_document
---
##load bugsigDB data##
```{r}
install.packages("devtools")
devtools::install_github("waldronlab/bugSigSimple", force = TRUE)
devtools::install_github("waldronlab/bugsigdbr", force = TRUE)
```

```{r}
library(bugSigSimple)
dat <- bugsigdbr::importBugSigDB(cache = FALSE) 
dim(dat)
names(dat)
```

```{r}
library(tidyverse)
library(Xmisc)
library(Hmisc)
condsnew <-c("COVID-19")
covid_gut <- subsetByCondition(dat,condsnew) %>%
    mutate(studyexp = paste(Study, Experiment, sep = "_")) %>%
    mutate(site = with(covid_gut, ifelse(`Body site` == "feces" | `Body site` == "rectal", "Gut", 
                       ifelse(`Body site` == "nasopharynx" | `Body site` == "nasopharynx,oropharynx", "aURT",`Body site`)))) %>%
    mutate(comparison1 = paste(`Group 0 name`, `Group 1 name`, sep = " vs "))
View(covid_gut)  
```  
  
##subset by bodysite
```{r}
library(dplyr)
gut_sigs <- subset(covid_gut, 
                           site == "Gut") %>%
    drop_na(Source)

naso_sigs <- subset(covid_gut, 
                           site == "aURT") %>%
  drop_na(Source)
View(gut_sigs)
```

##gut microbiota analysis##
##look specifically at case-control comparisons
```{r}
library(dplyr)
healthy <- grepl(".*(healthy.*COVID|COVID.*healthy).*", gut_sigs$comparison1, ignore.case = TRUE)

cc_gut_sigs1 <- gut_sigs %>%
      filter(healthy == TRUE) 
view(cc_gut_sigs1)
```

## Overall frequencies of taxa increased in cases for case/control feces studies
```{r}
View(getMostFrequentTaxa(cc_gut_sigs1, n=40))

View(getMostFrequentTaxa(cc_gut_sigs1,sig.type="increased"))
```

##binomial tests - significantly increased taxa in case-control comparisons
```{r}

increasedtaxa <- c("Enterococcus", "Streptococcus", "Rothia", "Enterococcaceae", "Lactobacillaceae", "Enterocloster citroniae", "Enterobacteriaceae", "Atopobium", "Lactobacillus", "Shigella")

freq <- c("8 / 8", "8 / 8", "5 / 5", "5 / 5", "5 / 8", "5 / 5", "5 / 5", "4 / 4", "4 / 4", "4 / 4")

binom.test(x=8, n=8) ##enterococcus
binom.test(x=8, n=8) ##streptococcus
binom.test(x=5, n=5) ##rothia
binom.test(x=5, n=5) ##enterococcaceae
binom.test(x=5, n=8) ##lactobacillaceae
binom.test(x=5, n=5) ##enterocloster citroniae
binom.test(x=5, n=5) ##enterobacteriaceae
binom.test(x=4, n=4)##atopobium
binom.test(x=4, n=4) ##lactobacillus
binom.test(x=4, n=4) ##shigella


## Adjusting p value by FDR correction for increased taxa

pvals <- c(0.007812, 0.007812, 0.0625,  0.0625, 0.7266,  0.0625,  0.0625, 0.125, 0.125, 0.125)
adj.pvals <- p.adjust(pvals, method="fdr")
adj.pvals

binom_inc <- data.frame(increasedtaxa, freq, pvals, adj.pvals)
View(binom_inc)
```

##output frequency/p-value table##
```{r}
library(gt)
library(tidyverse)
library(glue)

inctable <- binom_inc %>%
  gt(rowname_col = "row", groupname_col = "group") %>%
  tab_header(
    title = "Increased Taxa by Binomial Test - Gut Signatures"
  ) %>%
  cols_label(
    increasedtaxa = "Taxon", 
    freq = "Frequency", 
    pvals = "P-value", 
    adj.pvals="Adj. p-value") %>%
  tab_style(
    style = list(
      cell_fill(color = "lightgreen"),
     cell_text(weight = "bold")),
      locations = cells_body(
      rows = adj.pvals < 0.05)
      )

gtsave(inctable, "increasedtaxatable_gut.png")
```
##Monte-Carlo simulation for increased abundance taxa
```{r}
library(bugSigSimple)
relevant.sigs <- extractBugSigs(cc_gut_sigs1)
gut.sigs.increased <- extractBugSigs(cc_gut_sigs1, direction = "increased")
(my.siglengths.inc <- sapply(gut.sigs.increased, length))

getCriticalN(relevant.sigs, my.siglengths.inc)
# Compare to observed - enterococcus & streptococcus are the only taxa that equal the critical limit
frequencySigs(gut.sigs.increased)
```
## Overall frequencies of taxa decreased in cases for case/control feces studies
```{r}
View(getMostFrequentTaxa(cc_gut_sigs1, n=40))

View(getMostFrequentTaxa(cc_gut_sigs1,sig.type="decreased"))
```

##binomial tests - significantly decreased taxa in case-control comparisons
```{r}

decreasedtaxa<- c("Coprococcus", "Anaerostipes", "Roseburia", "Faecalibacterium", "Lachnospira", "Clostridium", "Dorea", "Anaerobutyricum hallii", "Coprococcus comes", "Dorea formicigenerans")
 
freq1 <- c("7 / 7", "7 / 8", "6 / 6", "6 / 7", "5 / 6", "4 / 6", "4 / 4", "4 / 6", "4 / 4", "4 / 4")

binom.test(x=7, n=7) ##coprococcus
binom.test(x=7, n=8) ##anaerostipes
binom.test(x=6, n=6) ##roseburia
binom.test(x=6, n=7) ##faecalibacterium
binom.test(x=5, n=6) ##lachnospira
binom.test(x=4, n=6) ##clostridium
binom.test(x=4, n=4) ##dorea
binom.test(x=4, n=6)##anaerobutyricum hallii
binom.test(x=4, n=4) ##coprococcus comes
binom.test(x=4, n=4) ##dorea formicigenerans

## Adjusting p value by FDR correction for increased taxa

pvals1 <- c(0.01563, 0.07031, 0.03125, 0.125,  0.2188, 0.6875,  0.125,  0.6875, 0.125, 0.125)
adj.pvals1 <- p.adjust(pvals1, method="fdr")
adj.pvals1

binom_dec <- data.frame(decreasedtaxa, freq1, pvals1, adj.pvals1)
View(binom_dec)

dectable <- binom_dec %>%
  gt(rowname_col = "row", groupname_col = "group") %>%
  tab_header(
    title = "Decreased Taxa by Binomial Test - Gut Signatures"
  ) %>%
  cols_label(
    decreasedtaxa = "Taxon", 
    freq1 = "Frequency", 
    pvals1 = "P-value", 
    adj.pvals1 ="Adj. p-value") %>%
  tab_style(
    style = list(
      cell_fill(color = "lightgreen"),
     cell_text(weight = "bold")),
      locations = cells_body(
      rows = adj.pvals1 < 0.05)
      )

gtsave(dectable, "decreasedtaxatable_gut.png")
```

##Monte-Carlo simulation for decreased abundance taxa
```{r}
library(bugSigSimple)
relevant.sigs <- extractBugSigs(cc_gut_sigs1)
gut.sigs.decreased <- extractBugSigs(cc_gut_sigs1, direction = "decreased")
(my.siglengths.inc <- sapply(gut.sigs.decreased, length))

getCriticalN(relevant.sigs, my.siglengths.inc)
# Compare to observed - no taxa reach critical limit
frequencySigs(gut.sigs.decreased)
```

##nasopharyngeal microbiota analysis##
##look specifically at case-control comparisons
```{r}
library(dplyr)
healthy <- grepl(".*(control.*COVID|COVID.*control).*", naso_sigs$comparison1, ignore.case = TRUE)

cc_naso_sigs1 <- naso_sigs %>%
      filter(healthy == TRUE) %>%
  subset(Study != "Study 458")
view(cc_naso_sigs1)
```

## Overall frequencies of taxa increased in cases for case/control URT studies
```{r}
View(getMostFrequentTaxa(cc_URT_sigs1, n=40))

View(getMostFrequentTaxa(cc_URT_sigs1,sig.type="increased"))
```

##binomial tests - significantly increased taxa in case-control comparisons
```{r}

increasedtaxa_naso <- c("Veillonella", "Rothia mucilaginosa", "Prevotella ruminicola", "Staphylococcaceae", "Granulicatella", "Enterococcaceae")

freq2 <- c("4 / 5", "2 / 2", "2 / 2", "2 / 2", "2 / 2", "2 / 2")

binom.test(x=4, n=5) ##veillonella
binom.test(x=2, n=2) ##rothia mucilaginosa
binom.test(x=2, n=2) ##prevotella ruminicola
binom.test(x=2, n=2) ##staphylococcaceae
binom.test(x=2, n=2) ##granulicatella
binom.test(x=2, n=2) ##enterococcaceae

## Adjusting p value by FDR correction for increased taxa

pvals2 <- c(0.375, 0.5, 0.5, 0.5, 0.5, 0.5)
adj.pvals2 <- p.adjust(pvals2, method="fdr")
adj.pvals2

binom_inc_naso <- data.frame(increasedtaxa_naso, freq2, pvals2, adj.pvals2)
View(binom_inc)

inctable.naso <- binom_inc_naso %>%
  gt(rowname_col = "row", groupname_col = "group") %>%
  tab_header(
    title = "Increased Taxa by Binomial Test - Nasopharyngeal Signatures"
  ) %>%
  cols_label(
    increasedtaxa_naso = "Taxon", 
    freq2 = "Frequency", 
    pvals2 = "P-value", 
    adj.pvals2 ="Adj. p-value") %>%
  tab_style(
    style = list(
      cell_fill(color = "lightgreen"),
     cell_text(weight = "bold")),
      locations = cells_body(
      rows = adj.pvals2 < 0.05)
      )

gtsave(inctable.naso, "increasedtaxatable_naso.png")
```

##Monte-Carlo simulation for increased abundance taxa
```{r}
library(bugSigSimple)
relevant.sigs <- extractBugSigs(cc_URT_sigs1)
URT.sigs.increased <- extractBugSigs(cc_URT_sigs1, direction = "increased")
(my.siglengths.inc <- sapply(URT.sigs.increased, length))

getCriticalN(relevant.sigs, my.siglengths.inc)
# Compare to observed - enterococcus & streptococcus are the only taxa that equal the critical limit
frequencySigs(URT.sigs.increased)
```

## Overall frequencies of taxa decreased in cases for case/control URT studies
```{r}
View(getMostFrequentTaxa(cc_URT_sigs1, n=40))

View(getMostFrequentTaxa(cc_URT_sigs1,sig.type="decreased"))
```

##binomial tests - significantly decreased taxa in case-control comparisons
```{r}

decreasedtaxa_naso <- c("Rothia", "Mobiluncus", "Bifidobacterium", "Corynebacterium", "Intrasporangiaceae", "Arthrobacter", "Kocuria", "Arachnia", "Rubrobacteraceae", "Rubrobacter")

freq3 <- c("3 / 3", "2 / 2", "2 / 2", "2 / 2", "2 / 2", "2 / 2", "2 / 2", "2 / 2", "2 / 2", "2 / 2")

binom.test(x=3, n=3) ##Rothia
binom.test(x=2, n=2) ##Mobiluncus
binom.test(x=2, n=2) ##Bifidobacterium
binom.test(x=2, n=2) ##Corynebacterium
binom.test(x=2, n=2) ##Intrasporangiaceae
binom.test(x=2, n=2) ##Arthrobacter
binom.test(x=2, n=2) ##Kocuria
binom.test(x=2, n=2)##Arachnia
binom.test(x=2, n=2) ##Rubrobacteraceae
binom.test(x=2, n=2) ##Rubrobacter

## Adjusting p value by FDR correction for increased taxa

pvals3 <- c(0.25, 0.5, 0.5, 0.5, 0.5, 0.5, 0.5, 0.5, 0.5, 0.5)
adj.pvals3 <- p.adjust(pvals3, method="fdr")
adj.pvals3

binom_dec_naso <- data.frame(decreasedtaxa_naso, freq3, pvals3, adj.pvals3)
View(binom_dec_naso)

dectable.naso <- binom_dec_naso %>%
  gt(rowname_col = "row", groupname_col = "group") %>%
  tab_header(
    title = "Decreased Taxa by Binomial Test - Nasopharyngeal Signatures"
  ) %>%
  cols_label(
    decreasedtaxa_naso = "Taxon", 
    freq3 = "Frequency", 
    pvals3 = "P-value", 
    adj.pvals3 ="Adj. p-value") %>%
  tab_style(
    style = list(
      cell_fill(color = "lightgreen"),
     cell_text(weight = "bold")),
      locations = cells_body(
      rows = adj.pvals3 < 0.05)
      )

gtsave(dectable.naso, "decreasedtaxatable_naso.png")
```

##Monte-Carlo simulation for decreased abundance taxa
```{r}
library(bugSigSimple)
relevant.sigs <- extractBugSigs(cc_URT_sigs1)
URT.sigs.decreased <- extractBugSigs(cc_URT_sigs1, direction = "decreased")
(my.siglengths.inc <- sapply(URT.sigs.decreased, length))

getCriticalN(relevant.sigs, my.siglengths.inc)
# Compare to observed 
frequencySigs(URT.sigs.decreased)
```
