# Setup phase
Install and load relevant packages
```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("waldronlab/bugSigSimple")
devtools::install_github("waldronlab/bugSigDB")
```

Load bugSigSimple and bugSigDB
```{r}
#library(bugSigDB)
library (bugSigSimple)
```

Read curation file
```{r}
dat <- bugsigdbr::importBugSigDB()
dim(dat)
```

#subsetting all signatures by curator and condition
```{r}
my.dat <- subsetByCurator(dat, curator="Fatima Zohra")
table(my.dat[,"Condition"])
condsnew <-c("bipolar disorder","unipolar depression")

my.dat.cond<- subsetByCondition(my.dat,condsnew)
table(my.dat.cond[,"Condition"])

```



#overall frequencies of taxa in up and down
```{r}
getMostFrequentTaxa(my.dat.cond,n=30)
getMostFrequentTaxa(my.dat.cond,sig.type="increased")
getMostFrequentTaxa(my.dat.cond,sig.type="decreased")
```


#binomial test in UP
```{r}
binom.test(x=4, n=6)
binom.test(x=4, n=4)
binom.test(x=4, n=4)
binom.test(x=4, n=5)
binom.test(x=4, n=5)
binom.test(x=3, n=4)
binom.test(x=3, n=5)
binom.test(x=3, n=3)
binom.test(x=3, n=3)
binom.test(x=3, n=4)
```
#binomial test in Down
```{r}
binom.test(x=7, n=8)
binom.test(x=6, n=6)
binom.test(x=4, n=5)
binom.test(x=4, n=5)
binom.test(x=3, n=5)
binom.test(x=3, n=5)
binom.test(x=2, n=6)
binom.test(x=2, n=2)
binom.test(x=2, n=3)
binom.test(x=2, n=2)
```

#adjusting p value by Bonferroni correction
```{r}
pvals <- c(0.6875, 0.125, 0.375, 0.625, 1, 0.25, 0.07031, 0.03125, 0.6875, 0.5)
adj.pvals <- p.adjust(pvals, method="bonferroni")
adj.pvals
```

#subsutting only on bipolar depression
```{r}
ind <- my.dat3[,"Condition"] %in% c("bipolar disorder")
my.dat.mdd <- my.dat3[!ind,]
dim(my.dat.mdd)
table(my.dat.mdd[,"Condition"])

getMostFrequentTaxa(my.dat.mdd, n=20)
getMostFrequentTaxa(my.dat.mdd,sig.type= "increased")
getMostFrequentTaxa(my.dat.mdd,sig.type= "decreased")
```
#binomial test in UP
```{r}
binom.test(x=3, n=3)
binom.test(x=3, n=3)
binom.test(x=3, n=4)
binom.test(x=2, n=4)
binom.test(x=2, n=3)
binom.test(x=2, n=2)
binom.test(x=2, n=4)
binom.test(x=2, n=3)
binom.test(x=2, n=4)
binom.test(x=2, n=2)
```
#binomial test in Down
```{r}
binom.test(x=4, n=5)
binom.test(x=4, n=4)
binom.test(x=4, n=5)
binom.test(x=3, n=5)
binom.test(x=2, n=4)
binom.test(x=2, n=4)
binom.test(x=2, n=4)
binom.test(x=2, n=4)
binom.test(x=2, n=2)
binom.test(x=2, n=2)
```

#adjusting p value by FDR correction for BD
```{r}
pvals <- c(0.6875, 0.125, 0.375, 0.625, 1, 0.25, 0.07031, 0.03125, 0.6875, 0.5)
adj.pvals <- p.adjust(pvals, method="fdr")
adj.pvals
```


#subsutting only on MDD 
```{r}
ind <- my.dat3[,"Condition"] %in% c("unipolar depression")
my.dat.bd <- my.dat3[!ind,]
dim(my.dat.bd)
table(my.dat.bd[,"Condition"])

getMostFrequentTaxa(my.dat.bd, n=20)
getMostFrequentTaxa(my.dat.bd,sig.type= "increased")
getMostFrequentTaxa(my.dat.bd,sig.type= "decreased")
```

#binomial test in UP w/out MHT
```{r}
binom.test(x=4, n=4)
binom.test(x=4, n=4)
binom.test(x=4, n=4)
binom.test(x=2, n=2)
binom.test(x=2, n=2)
binom.test(x=2, n=2)
binom.test(x=2, n=3)
binom.test(x=1, n=2)
```


#binomial test in Down w/out MHT
```{r}
binom.test(x=4, n=5)
binom.test(x=4, n=4)
binom.test(x=4, n=5)
binom.test(x=3, n=5)
binom.test(x=2, n=4)
binom.test(x=2, n=4)
binom.test(x=2, n=4)
binom.test(x=2, n=4)
binom.test(x=2, n=2)
binom.test(x=2, n=2)
```
#faecalibacterium plot
```{r}
taxa.mdd <- c(up=1/7, down=4/7)
taxa.bd <- c(up=0/4, down=3/4)
barplot(rbind(taxa.mdd, taxa.bd), beside=TRUE, col=c("red","blue"), legend=TRUE, args.legend=list(x="topleft", legend=c("major depressive disorder", "bipolar disorder")), ylab="relative frequency", main="Frequency of Faecalibacterium")
```
 
#Actinobacteria plot
```{r}
taxa.mdd <- c(up=2/7, down=2/7)
taxa.bd <- c(up=2/4, down=0/4)
barplot(rbind(taxa.mdd, taxa.bd), beside=TRUE, col=c("red","blue"), legend=TRUE, args.legend=list(x="topright", cex = 0.70, legend=c("major depressive disorder", "bipolar disorder")), ylab="relative frequency", main="Frequency of Actinobacteria")
```

#Bacteroides plot
```{r}

taxa.mdd <- c(up=2/7, down=2/7)
taxa.bd <- c(up=0/4, down=2/4)
barplot(rbind(taxa.mdd, taxa.bd), beside=TRUE, col=c("red","blue"), legend=TRUE, args.legend=list(x="topleft", cex = 0.80, legend=c("major depressive disorder", "bipolar disorder")), ylab="relative frequency", main="Frequency of Bacteroides")
```

#Ruminococcaceae plot
```{r}
taxa.mdd <- c(up=0/7, down=4/7)
taxa.bd <- c(up=0/4, down=2/4)
barplot(rbind(taxa.mdd, taxa.bd), beside=TRUE, col=c("red","blue"), legend=TRUE, args.legend=list(x="topleft", cex = 0.75, legend=c("major depressive disorder", "bipolar disorder")), ylab="relative frequency", main="Frequency of Ruminococcaceae")
```






