---
title: "The Irritable Bowel Syndrome-associated Microbiome"
author: "Kweku Amoo"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Irritable Bowel Syndrome-associated Microbiome}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Analysis objectives

1. Import, recode, and subset data from bugsigdb.org    
2. Create a table of studies
3. Create a clustered heatmap showing similarity of signatures from independent studies
4. Calculate the frequency of appearance of each taxa in independent signatures, and identify the most frequently reported taxa
5. Estimate the probability of the most frequently identified taxa occuring so frequently by chance

# Packages installation

Install packages (not evaluated in vignette)

```{r, eval=FALSE}
install.packages(c("devtools", "tidyverse", "kableExtra", "gt", "glue"))
devtools::install_github("waldronlab/bugSigSimple")
devtools::install_github("waldronlab/BugSigDBStats")
devtools::install_github("waldronlab/bugsigdbr")
```

# Data import, recoding, and subset

```{r}
library(bugSigSimple)
dat <- bugsigdbr::importBugSigDB(cache = FALSE) 
dim(dat)
names(dat)
```

```{r}
library(tidyverse)
condition_of_interest <- c("irritable bowel syndrome")
efo <- bugsigdbr::getOntology("efo")
dat_condition <- bugsigdbr::subsetByOntology(dat, column = "Condition", "irritable bowel syndrome", efo) %>%
  mutate(comparison1 = paste(`Group 0 name`, `Group 1 name`, sep = " vs "))
```  
  

## Table of studies

```{r}
bugSigSimple::createStudyTable(dat_condition)
```


## Taxon frequency tables by body site

```{r}
gut_sigs <- filter(dat_condition,
                   `Body site` %in% c("Feces, Mucosa of small intestine", "Feces"))
```

In this table, the Binomial Test p-value corresponds to the null hypothesis

H0: the proportion of signatures in which the taxon is reported increased or decreased, relative to the total number of signatures in which it is reported, is equal to 0.5

```{r}
kableExtra::kbl(bugSigSimple::createTaxonTable(gut_sigs))
```

# Cluster analysis

Note, this EDA should really be done _before_ hypothesis testing. 

First calculate pairwise overlaps for all signatures of length > 1:
```{r}
allsigs <- bugsigdbr::getSignatures(dat_condition, tax.id.type = "taxname")
allsigs <- allsigs[sapply(allsigs, length) > 1] #require length > 1
length(allsigs)
mydists <- BugSigDBStats::calcPairwiseOverlaps(allsigs)
dim(mydists)
```
What is the distribution of signature lengths?
```{r}
library(ggplot2)
siglengths <- sapply(allsigs, length)
siglengths.df <- data.frame(siglengths = siglengths)
ggplot(siglengths.df, aes(x=siglengths)) +
  geom_bar()
```
```{r}
table(siglengths)
```

Create a matrix of Jaccard similarities (0 for no overlap, 1 for 100% overlap)

```{r}
jmat <- BugSigDBStats::calcJaccardSimilarity(allsigs)
```

```{r, fig.width=20, fig.height=20}
library(ComplexHeatmap)
ha <- HeatmapAnnotation(`Signature Length` = anno_barplot(siglengths))
hr <- rowAnnotation(`Signature Length` = anno_barplot(siglengths))
hm <- Heatmap(
  jmat,
  top_annotation = ha, left_annotation = hr,
  row_names_max_width = unit(20, "cm"),
  column_names_max_height = unit(20, "cm"),
#  row_labels = sub(".+:", "", rownames(jmat)),  #get rid of study labels
  column_labels = sub(".+:", "", colnames(jmat))
)
hm
```

Use this interactively to make an interactive heatmap. Some expanding of the default size is required to see anything. Creating a sub-heatmap, then exporting it as a table, allows in-depth identification of the subgroups.

```{r, eval = FALSE}
library(InteractiveComplexHeatmap)
hm <- draw(hm)
htShiny(hm)
```

```{r, fig.height=60, fig.width=20}
hc <- hclust(dist(jmat))
plot(hc)
```

This tree can be cut to show the clusters, for example. The clusters of more than 1 signature but less than ~10 signatures are most likely to be something interesting.

```{r}
clusts <- sort(cutree(hc, k = 8))  #set the number of clusters here with k
lapply(unique(clusts), function(i) names(clusts)[clusts == i])
```

# Create a wide-format dataframe

This would be suitable for regression analysis.

```{r}
dat_withsigs <- filter(dat_condition, !is.na(dat_condition$`NCBI Taxonomy IDs`))
sigs <- bugsigdbr::getSignatures(dat_withsigs, tax.id.type = "taxname")
cmat <- t(safe::getCmatrix(sigs, as.matrix = TRUE, min.size = 0, prune = FALSE))
cdf <- data.frame(cmat, stringsAsFactors = FALSE, check.names = FALSE)
cdf <- cbind(dat_withsigs, cdf)
colnames(cdf)[1:54]
```

Note this has a number of columns that are mostly zeros, it could be filtered significantly for any regression or machine learning analysis:

```{r}
table(cdf[["Bifidobacterium catenulatum"]])
```

Create another heatmap on correlations of presence/absence of taxa. This is not necessary because the previous Jaccard Index heatmap is probably better, it is just a demonstration of doing something with the taxa presence/absence directly. 

```{r, fig.width=20, fig.height=20}
sigcors <- cor(t(cmat))
siglengths <- sapply(sigs, length)
ha <- HeatmapAnnotation(`Signature Length` = anno_barplot(siglengths))
hr <- rowAnnotation(`Signature Length` = anno_barplot(siglengths))
hm <- Heatmap(
  sigcors,
  top_annotation = ha, left_annotation = hr,
  row_names_max_width = unit(20, "cm"),
  column_names_max_height = unit(20, "cm"),
 # row_labels = sub(".+:", "", rownames(sigcors)), ##removing study just to make signature names legible
  column_labels = sub(".+:", "", colnames(sigcors))
)
hm
```

Use this interactively to make an interactive heatmap:
```{r, eval = FALSE}
library(InteractiveComplexHeatmap)
hm <- draw(hm)
htShiny(hm)
```

