---
title: "Capstone Analysis Samara Khan"
output: html_document
---

# Analysis objectives

1. Import, recode, and subset data from bugsigdb.org    
2. Create a table of studies
3. Calculate the frequency of appearance of each taxa in independent signatures and identify the most frequently reported taxa
4. Conduct a Monte Carlo simulation to test significance of frequency of most recurrent taxa
5. Fit a log-linear model to standardized HMP data 

#Install packages
```{r setup, eval=FALSE, messages=FALSE}
install.packages(c("devtools", "tidyverse", "kableExtra", "gt", "glue"))
devtools::install_github("waldronlab/bugSigSimple")
devtools::install_github("waldronlab/BugSigDBStats")
devtools::install_github("waldronlab/bugsigdbr")
```

#Load packages and subset data
```{r}
library(bugSigSimple)
library(BugSigDBStats)
library(bugsigdbr)
library(tidyverse)
library(devtools)
library(kableExtra)
library(dplyr)
library(gt)
library(glue)
```


#Loading and subsetting data to include only signatures from the female reproductive tract

```{r}
dat <- bugsigdbr::importBugSigDB(cache= TRUE)
dim(dat)
names(dat)

subset.dat <-
  dat[which(
    dat$PMID == "30778155" |
      dat$PMID == "32192080" |
      dat$PMID == "31087436" |
      dat$PMID == "26901400" |
      dat$PMID == "33839907" |
      dat$PMID == "32046455" |
      dat$PMID == "33925708" |
      dat$PMID == "32299442" |
      dat$PMID == "33313185" |
      dat$PMID == "34268384" | dat$PMID == "33660232"
  ), ]

subset.dat2 <-
  dat[which(
    dat$PMID == "30778155" |
      dat$PMID == "32192080" |
      dat$PMID == "31087436" |
      dat$PMID == "26901400" |
      dat$PMID == "32046455" |
      dat$PMID == "33925708" |
      dat$PMID == "32299442" |
      dat$PMID == "33313185" | dat$PMID == "34268384"
  ), ]

reproductive_sigs <-
  subset.dat2[which(subset.dat2$`Body site` != "feces" |
                      is.na(subset.dat2$`Body site`)), ]

```

#Create table of studies
```{r}
bugSigSimple::createStudyTable(reproductive_sigs)

kableExtra::kbl(bugSigSimple::createTaxonTable(reproductive_sigs))
```

#Monte Carlo simulation for increased and decreased taxa
```{r}
sigs.increased <- filter(reproductive_sigs, `Abundance in Group 1` == "increased") %>% 
  bugsigdbr::getSignatures(tax.id.type = "taxname")

my.siglengths.inc <- sapply(sigs.increased, length)

getCriticalN(sigs.increased, my.siglengths.inc, alpha = 0.05, nsim = 1000)
frequencySigs(sigs.increased)

sigs.decreased <- filter(reproductive_sigs, `Abundance in Group 1` == "decreased") %>% 
  bugsigdbr::getSignatures(tax.id.type = "taxname")

my.siglengths.dec <- sapply(sigs.decreased, length)

getCriticalN(sigs.decreased, my.siglengths.dec, alpha = 0.05, nsim = 1000)
frequencySigs(sigs.decreased)

```

#Cluster analysis
```{r}
clustersigs <- bugsigdbr::getSignatures(reproductive_sigs, tax.id.type = "taxname")
clustersigs <- clustersigs[sapply(clustersigs, length) > 1]
dim(clustersigs)

mydists <- BugSigDBStats::calcPairwiseOverlaps(clustersigs)
dim(mydists)
```

#Create a matrix of Jaccard similarities (0 for no overlap, 1 for 100% overlap)

```{r}
signames <- unique(c(mydists$name1, mydists$name2))
jmat <- matrix(NA, nrow=length(signames), ncol=length(signames), dimnames=list(signames, signames))
diag(jmat) <- 1
for (i in 1:nrow(mydists)){
  jmat[mydists[i, "name2"], mydists[i, "name1"]] <- mydists[i, "jaccard"]
  jmat[mydists[i, "name1"], mydists[i, "name2"]] <- mydists[i, "jaccard"]
}
```

#Create a heatmap

```{r}
library(ComplexHeatmap)
siglengths <- sapply(reproductive_sigs, length)
ha <- HeatmapAnnotation(`Signature Length` = anno_barplot(siglengths))
hr <- rowAnnotation(`Signature Length` = anno_barplot(siglengths))
hm <- Heatmap(
  jmat,
  top_annotation = ha, left_annotation = hr,
  row_names_max_width = unit(20, "cm"),
  column_names_max_height = unit(20, "cm"),
  row_labels = sub(".+:", "", rownames(jmat)),
  column_labels = sub(".+:", "", colnames(jmat))
)
hm
```

#Create a data frame for regression analysis from healthy controls

```{r}
BiocManager::install("curatedMetagenomicData")
install.packages("purrr")

library(curatedMetagenomicData)
library(dplyr)
library(DT)
library(purrr)

taxon_strep <- c("1884" = "Streptococcus")
taxon_lact <- c("1578" = "Lactobacillus")

target_sample <- "vagina"
sample_metadata <- sampleMetadata %>% 
    filter(
        grepl(target_sample, body_site) | grepl(target_sample, body_subsite)
    ) %>% 
    discard(~all(is.na(.x))) %>% 
    as_tibble()
dim(sample_metadata)
unique(sample_metadata$study_name)

taxa_data <- rowData(tse) %>% 
    as.data.frame() %>% 
    tibble::rownames_to_column(var = "tax_name")

filtered_taxa <- taxa_data %>% 
    filter(
        tax_name == taxon_strep |
            genus == taxon_strep |
            genus == taxon_lact
    )

select_taxa <- filtered_taxa$tax_name

tidy_abundance <- assay(filtered_taxa) %>% 
    t() %>% 
    as.data.frame() %>% 
    tibble::rownames_to_column("SAMPLE") %>% 
    as_tibble() %>% 
    pivot_longer(
        names_to = "GENUS", values_to = "ABUNDANCE", 
        cols = 2:last_col()
    ) %>% 
    left_join(sample_metadata, by = c("SAMPLE" = "sample_id"))

head(tidy_abundance)

```

#Examining and cleaning data for log-linear model

```{r}

str(tidy_abundance)

nrow(data[is.na(tidy_abundance$age) | is.na(tidy_abundance$BMI),])
data[is.na(tidy_abundance$age) | is.na(tidy_abundance$BMI),]



```

#Poisson regression
```{r}
poisson <- glm(taxon_lact ~ taxon_strep + study_name + age + BMI)
summary (poisson)
```

Need to check for over/ under-dispersion (residual deviance >> df in model output)

