---
title: "Delivery Mode and the Meconium Microbiome"
name: "Shaimaa Elsafoury"
date: "2/24/2022"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Endometriosis-associated Microbiome}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Analysis objectives

1. Import, recode, and subset data from bugsigdb.org    
2. Create a table of studies
3. Calculate the frequency of appearance of each taxa in independent signatures and identify the most frequently reported taxa

# Making sure packages are installed

```{r, eval=FALSE, messages=FALSE}
install.packages(c("devtools", "tidyverse", "kableExtra"))
devtools::install_github("waldronlab/bugSigSimple")
devtools::install_github("waldronlab/BugSigDBStats")
devtools::install_github("waldronlab/bugsigdbr")
```

# Load and subset data

```{r, messages=FALSE}
library(bugSigSimple)
library(BugSigDBStats)
library(bugsigdbr)
library(tidyverse)
library(stringr)
library(kableExtra)
library(dplyr)
```

```{r}
# use version="devel" and cache = FALSE to take the latest version from bugsigdb.org
dat <- bugsigdbr::importBugSigDB(version = "devel", cache = TRUE) 
dim(dat)
names(dat)
```

# Subsetting

Subsetting only studies done on humans - 6 of 17 studies were excluded because they were either done on mice or not statistically reliable (Cregger et. al)

```{r, messages=FALSE}
subset.dat <-
  dat[which(
    dat$PMID == "30778155" |
      dat$PMID == "32192080" |
      dat$PMID == "31087436" |
      dat$PMID == "26901400" |
      dat$PMID == "33839907" |
      dat$PMID == "32046455" |
      dat$PMID == "33925708" |
      dat$PMID == "32299442" |
      dat$PMID == "33313185" |
      dat$PMID == "34268384" | dat$PMID == "33660232"
  ), ]
```

# All studies

Summary of studies and most frequent taxa increased and decreased in endometriosis patients for all studies

```{r, messages=FALSE}
bugSigSimple::createStudyTable(subset.dat)
```

```{r}
getMostFrequentTaxa(subset.dat,n=10)
```

```{r}
getMostFrequentTaxa(subset.dat,sig.type="increased")
```

```{r}
getMostFrequentTaxa(subset.dat,sig.type="decreased")
```
