---
title: "feces sigcomp"
author: "clare"
date: "8/2/2021"
output: html_document
---

# Analysis objectives

1. Import, recode, and subset data from bugsigdb.org    
2. Create a table of studies
3. Create a clustered heatmap showing similarity of signatures from independent studies
4. Calculate the frequency of appearance of each taxa in independent signatures, and identify the most frequently reported taxa
5. Estimate the probability of the most frequently identified taxa occuring so frequently by chance

# Packages installation

Install packages (not evaluated in vignette)

```{r, eval=FALSE}
install.packages(c("devtools", "tidyverse", "kableExtra", "gt", "glue"))
devtools::install_github("waldronlab/bugSigSimple")
devtools::install_github("waldronlab/BugSigDBStats")
devtools::install_github("waldronlab/bugsigdbr")
```

# Data import, recoding, and subset

```{r}
library(bugSigSimple)
dat <- bugsigdbr::importBugSigDB(cache = TRUE) 
dim(dat)
names(dat)
```

```{r}
library(tidyverse)
condsnew <- c("COVID-19")
covid_all <- subsetByCondition(dat, condsnew) %>%
  mutate(studyexp = paste(Study, Experiment, sep = "_")) %>%
  mutate(
    site = recode(`Body site`,
      "feces" = "Gut",
      "rectal" = "Gut",
      "nasopharynx" = "aURT",
      "oropharynx" = "aURT",
      "nasopharynx,oropharynx" = "aURT",
      "tongue" = "aURT"
    )
  ) %>%
  mutate(comparison1 = paste(`Group 0 name`, `Group 1 name`, sep = " vs "))
```  
  

## Table of studies

```{r}
bugSigSimple::createStudyTable(covid_all)
```


## Taxon frequency tables by body site

```{r}
library(dplyr)
gut_sigs <- filter(covid_all, 
                           site == "Gut") %>%
    drop_na(Source)

naso_sigs <- filter(covid_all, 
                           site == "aURT") %>%
  drop_na(Source)
```

In this table, the Binomial Test p-value corresponds to the null hypothesis

H0: the proportion of signatures in which the taxon is reported increased or decreased, relative to the total number of signatures in which it is reported, is equal to 0.5

```{r}
kableExtra::kbl(bugSigSimple::createTaxonTable(gut_sigs))
```

```{r}
kableExtra::kbl(bugSigSimple::createTaxonTable(naso_sigs))
```

# gut microbiota analysis

Look specifically at case-control comparisons
```{r}
healthy <- grepl(".*(healthy.*COVID|COVID.*healthy).*", gut_sigs$comparison1, ignore.case = TRUE)

cc_gut_sigs1 <- gut_sigs %>%
      filter(healthy == TRUE) 
cc_gut_sigs1
```

## Overall frequencies of taxa increased in cases for case/control feces studies

Identifying a taxon reported consistently in 8 out of 8 studies is _much_ more compelling that the FDR value here would suggest, since this taxon also passed a significance threshold in every one of those studies.

```{r}
cc_gut_sigs1_taxontable <- bugSigSimple::createTaxonTable(cc_gut_sigs1) %>% 
  mutate(FDR =  p.adjust(p = `Binomial Test pval`, method="fdr")) %>%
  relocate(FDR, .after = `Binomial Test pval`)
cc_gut_sigs1_taxontable
```




## Monte-Carlo simulation for increased abundance taxa

Just for the increased cc_gut_sigs1 for now. I am inclined to skip this analysis in favor of the clustering and binomial test analysis.

```{r}
library(bugSigSimple)
gut.sigs.increased <- filter(cc_gut_sigs1, `Abundance in Group 1` == "increased") %>% 
  bugsigdbr::getSignatures(tax.id.type = "taxname")
 my.siglengths.inc <- sapply(gut.sigs.increased, length)

getCriticalN(gut.sigs.increased, my.siglengths.inc)
# Compare to observed - enterococcus & streptococcus are the only taxa that equal the critical limit
frequencySigs(gut.sigs.increased)
```
## Overall frequencies of taxa decreased in cases for case/control feces studies
```{r, eval = FALSE}
createTaxonTable(cc_gut_sigs1, n=40)
```



## nasopharyngeal microbiota analysis

Look specifically at case-control comparisons

```{r}
library(dplyr)
healthy <- grepl(".*(control.*COVID|COVID.*control).*", naso_sigs$comparison1, ignore.case = TRUE)

cc_naso_sigs1 <- naso_sigs %>%
      filter(healthy == TRUE) %>%
  subset(Study != "Study 458")
cc_naso_sigs1
```

