---
title: "Fieldwork Analysis"
name: "Samara Khan"
output: html_document
---

# Analysis objectives

1. Import, recode, and subset data from bugsigdb.org    
2. Create a table of studies
3. Calculate the frequency of appearance of each taxa in independent signatures and identify the most frequently reported taxa
4. Estimate the probability of the most frequently identified taxa occuring by chance

# Making sure packages are installed

```{r, eval=FALSE, messages=FALSE}
install.packages(c("devtools", "tidyverse", "kableExtra"))
devtools::install_github("waldronlab/bugSigSimple")
devtools::install_github("waldronlab/BugSigDBStats")
devtools::install_github("waldronlab/bugsigdbr")
```

# Load and subset data

```{r, messages=FALSE}
library(bugSigSimple)
library(BugSigDBStats)
library(bugsigdbr)
library(tidyverse)
library(stringr)
library(kableExtra)
library(dplyr)
```

```{r}
dat <- bugsigdbr::importBugSigDB(cache= TRUE)
dim(dat)
names(dat)
```

# Subsetting

Subsetting only studies done on humans - 6 of 17 studies were excluded because they were either done on mice or not statistically reliable (Cregger et. al)

```{r, messages=FALSE}
subset.dat <-
  dat[which(
    dat$PMID == "30778155" |
      dat$PMID == "32192080" |
      dat$PMID == "31087436" |
      dat$PMID == "26901400" |
      dat$PMID == "33839907" |
      dat$PMID == "32046455" |
      dat$PMID == "33925708" |
      dat$PMID == "32299442" |
      dat$PMID == "33313185" |
      dat$PMID == "34268384" | dat$PMID == "33660232"
  ), ]
```

# All studies

Summary of studies and most frequent taxa increased and decreased in endometriosis patients for all studies

```{r, messages=FALSE}
bugSigSimple::createStudyTable(subset.dat)

getMostFrequentTaxa(subset.dat,n=30)
getMostFrequentTaxa(subset.dat,sig.type="increased")
getMostFrequentTaxa(subset.dat,sig.type="decreased")

```

# Excluding feces samples

Summary of studies and most frequent taxa in only samples from female reproductive tract, excluding feces samples

```{r, messages=FALSE}
subset.dat2 <-
  dat[which(
    dat$PMID == "30778155" |
      dat$PMID == "32192080" |
      dat$PMID == "31087436" |
      dat$PMID == "26901400" |
      dat$PMID == "32046455" |
      dat$PMID == "33925708" |
      dat$PMID == "32299442" |
      dat$PMID == "33313185" | dat$PMID == "34268384"
  ), ]

reproductive_sigs <-
  subset.dat2[which(subset.dat2$`Body site` != "feces" |
                      is.na(subset.dat2$`Body site`)), ]

bugSigSimple::createStudyTable(reproductive_sigs)

allfreqs <- bugSigSimple::createTaxonTable(reproductive_sigs, n = 20) #%>% #could change number
  arrange(I(decreased_signatures - increased_signatures))
incfreqs <- filter(allfreqs, I(increased_signatures - decreased_signatures) > 0)
decfreqs <- filter(allfreqs, I(increased_signatures - decreased_signatures) < 0)
kableExtra::kbl(allfreqs) %>%
  kable_paper("hover", full_width = FALSE)
```

## Load HMP data from curatedMetagenomicData

```{r, message=FALSE}
suppressPackageStartupMessages(library(curatedMetagenomicData))
# hmp_vagina_metadata <- filter(sampleMetadata, study_name == "HMP_2012" & body_site == "vagina")
all_healthy_adult_vagina_metadata <- filter(sampleMetadata, body_site == "vagina" & disease == "healthy" & age_category == "adult")
se <-
  curatedMetagenomicData::returnSamples(all_healthy_adult_vagina_metadata, dataType = "relative_abundance")
```

Get matrices of species and genus relative abundance
```{r}
allranks <- mia::splitByRanks(se)
species_relab <- t(assay(allranks[["species"]]))
genus_relab <- t(assay(allranks[["genus"]]))
family_relab <- t(assay(allranks[["family"]]))
```

Calculate prevalences at genus level
```{r}
genus_prev <- apply(genus_relab, 2, function(x) sum(x > 0) / length(x))
```

```{r}
genus_prev[names(genus_prev) %in% incfreqs$`Taxon Name`] #prevalence of genera increased in cases
genus_prev[names(genus_prev) %in% decfreqs$`Taxon Name`] #prevalence of genera decreased in cases
```


# Summary of studies and most frequent taxa in only fecal samples

```{r}

subset.dat3 <-dat[which(dat$PMID=="33839907" | dat$PMID=="33660232"| dat$PMID=="30778155" ),]
subset.dat4 <-subset.dat3[which(subset.dat3$`Body site`!="vagina"),]
feces_sigs<-subset.dat4[which(subset.dat4$`Body site`!="uterine cervix"),]

bugSigSimple::createStudyTable(feces_sigs)
getMostFrequentTaxa(feces_sigs,n=30)
getMostFrequentTaxa(feces_sigs,sig.type="increased")
getMostFrequentTaxa(feces_sigs,sig.type="decreased")

kableExtra::kbl(bugSigSimple::createTaxonTable(feces_sigs)) %>%
kable_paper("hover", full_width = F)

```



