---
title: "Analysis of prevalent stool genera in bugsigdb.org"
name: "Haoyan Zhong"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysis of prevalent stool genera in bugsigdb.org}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## BugsigDB stat- Signature overlaps

Compare genus abundance change (increased/decreased) from all diseased subjects in BugsigDB with existing typical genus from healthy volunteers recorded in BugsigDB Study 562
typical genus were set at prevalence level=0, 50, and 70
Using Feces sample as an example

## Get typical healthy signatures from Study 562

```{r healthy50}
library(dplyr)
library(bugsigdbr)
library(BugSigDBStats)
library(ggpubr)

full.dat <- bugsigdbr::importBugSigDB(version = "devel", cache=FALSE)
#Stripping empty signatures:
ind1 <- lengths(full.dat[["MetaPhlAn taxon names"]]) > 0
ind2 <- lengths(full.dat[["NCBI Taxonomy IDs"]]) > 0
dat <- full.dat[ind1 & ind2,]

#prevalence threshold=50%
bugs50<- (dat) %>%
  filter (Study=='Study 562') %>%
  filter (Experiment=="Experiment 3" & `Signature page name`=='Signature 1') %>%
  filter (`Body site` == "Feces") %>%
  filter (grepl("genus",Description))
healthysig50 <- bugsigdbr::getSignatures(bugs50)  
head(healthysig50)
```

# this code didn't work because experiment 10 has no signatures
#prevalence threshold=0

```
bugs0<- (dat) %>%
  filter (Study=='Study 562') %>%
  filter (Experiment=="Experiment 10") %>%
  filter (`Body site` == "Feces") %>%
  filter (grepl("genus",Description))
healthysig0 <- bugsigdbr::getSignatures(bugs0)  
head(healthysig0)
```


```{r healthy70}
#prevalence threshold=70%
bugs70<- (dat) %>%
  filter (Study=='Study 562') %>%
  filter (Experiment=="Experiment 9" & `Signature page name`=='Signature 1') %>%
  filter (`Body site` == "Feces") %>%
  filter (grepl("genus",Description))
healthysig70 <- bugsigdbr::getSignatures(bugs70)  
head(healthysig70)
```

## Get changed in abundance disease genus
```{r increased disease genus}
bugdisease_increased <-  dat %>%
  filter (`Host species` == 'Homo sapiens') %>%
  filter (`Body site` == 'Feces') %>%
  filter (`Abundance in Group 1`=="increased") 
#get genus level sig change in abundance only
diseasesig_increased <- bugsigdbr::getSignatures(bugdisease_increased, tax.level='genus')
length(diseasesig_increased)

regexp <- "[[:digit:]]+"
bugdisease_increased2<-bugdisease_increased%>%mutate(
                     study_id=stringr::str_extract(Study, regexp),
                     experiment_id=stringr::str_extract(Experiment, regexp),
                     signature_id=stringr::str_extract(`Signature page name`, regexp)
                     )

diseasesig_increased2 <- purrr::map_df(diseasesig_increased, ~as.data.frame(.x), .id="id") %>%
#  tidyr::pivot_wider(id_cols=id) %>%
  mutate(name2_1=stringr::str_replace(id, "bsdb:", "")) %>%
  mutate(name2_2=(stringr::str_split(name2_1,"/"))) %>%
  mutate(study_id= purrr::map_chr(name2_2, 1),
         experiment_id= purrr::map_chr(name2_2, 2),
         signature_id= purrr::map_chr(stringr::str_split(purrr::map_chr(name2_2, 3),'_'),1)
  ) %>%
  dplyr::select(-c("name2_1","name2_2"))

#exclude studies does not return genus level abundance change
bugall_increased <- merge(x = bugdisease_increased2, y = diseasesig_increased2, 
                          by =c('study_id','experiment_id','signature_id'), all.x = TRUE)
```
```{r decreased disease genus}
bugdisease_decreased <-  dat %>%
  filter (`Host species` == 'Homo sapiens') %>%
  filter (`Body site` == 'Feces') %>%
  filter (`Abundance in Group 1`=="decreased") 
#get genus level sig change in abundance only
diseasesig_decreased <- bugsigdbr::getSignatures(bugdisease_decreased, tax.level='genus')
length(diseasesig_decreased)

regexp <- "[[:digit:]]+"
bugdisease_decreased2<-bugdisease_decreased%>%mutate(
                     study_id=stringr::str_extract(Study, regexp),
                     experiment_id=stringr::str_extract(Experiment, regexp),
                     signature_id=stringr::str_extract(`Signature page name`, regexp)
                     )

diseasesig_decreased2 <- purrr::map_df(diseasesig_decreased, ~as.data.frame(.x), .id="id") %>%
#  tidyr::pivot_wider(id_cols=id) %>%
  mutate(name2_1=stringr::str_replace(id, "bsdb:", "")) %>%
  mutate(name2_2=(stringr::str_split(name2_1,"/"))) %>%
  mutate(study_id= purrr::map_chr(name2_2, 1),
         experiment_id= purrr::map_chr(name2_2, 2),
         signature_id= purrr::map_chr(stringr::str_split(purrr::map_chr(name2_2, 3),'_'),1)
  ) %>%
  dplyr::select(-c("name2_1","name2_2"))

#exclude studies does not return genus level abundance change
bugall_decreased <- merge(x = diseasesig_decreased2, y = bugdisease_decreased2, 
                          by =c('study_id','experiment_id','signature_id'), all.x = TRUE)
```

## Calculate pairwise overlaps

#this code didn't work because experiment 10 has no signatures, therefore line 130 - 180 were commented out ```{r overlap_prevelance=0}
#```{r overlap_prevelance=0} experiment 10 has no signatures

library(purrr)
list0_decreased <- c(healthysig0, diseasesig_decreased)
paircomp0_decreased <- calcPairwiseOverlaps(list0_decreased)

paircheck0_decreased <- paircomp0_decreased %>%
  filter(stringr::str_detect(name1, 'feces-from-healthy-adult_vs_none---0%-prevalence-threshold')) %>%
  mutate(name2_1=stringr::str_replace(name2, "bsdb:", "")) %>%
  mutate(name2_2=(stringr::str_split(name2_1,"/"))) %>%
  mutate(study_id= purrr::map_chr(name2_2, 1),
          experiment_id= purrr::map_chr(name2_2, 2),
          signature_id= purrr::map_chr(stringr::str_split(purrr::map_chr(name2_2, 3),'_'),1)
        ) 

list0_increased <- c(healthysig0, diseasesig_increased)
paircomp0_increased <- calcPairwiseOverlaps(list0_increased)

paircheck0_increased <- paircomp0_increased %>%
  filter(stringr::str_detect(name1, 'feces-from-healthy-adult_vs_none---0%-prevalence-threshold')) %>%
  mutate(name2_1=stringr::str_replace(name2, "bsdb:", "")) %>%
  mutate(name2_2=(stringr::str_split(name2_1,"/"))) %>%
  mutate(study_id= purrr::map_chr(name2_2, 1),
          experiment_id= purrr::map_chr(name2_2, 2),
          signature_id= purrr::map_chr(stringr::str_split(purrr::map_chr(name2_2, 3),'_'),1)
        ) 
```
# merge jacard scores with disease study data
The issue why I need to merge overlap score back with all disease budsigdb studies is that calcPairwiseOverlaps drops all 0 intersect pairs

#```{r merge overlap scores with all disease study data}
overlap0_decreased<- merge(x = bugall_decreased, y = paircheck0_decreased, by =c('study_id','experiment_id','signature_id'), all.x = TRUE) %>%
  mutate(type="decreased")%>%
  filter(length_set2>4) #keep only bugsigdb studies that tested more than 4 increased or decreased
overlap0_decreased[is.na(overlap0_decreased)] <- 0  

overlap0_increased<- merge(x = bugall_increased, y = paircheck0_increased, by =c('study_id','experiment_id','signature_id'), all.x = TRUE) %>%
  mutate(type="increased") %>%
  filter(length_set2 > 4) #keep only bugsigdb studies that tested more than 4 increased or decreased
overlap0_increased[is.na(overlap0_increased)] <- 0  
```

#```{r comparison-increased_vs_decrease}
```
t.test(overlap0_increased$overlap, overlap0_decreased$overlap)
wilcox.test(overlap0_increased$overlap, overlap0_decreased$overlap)

p <- rbind(overlap0_increased,overlap0_decreased) %>%
  ggplot( aes(x=overlap, fill=type)) +
    geom_histogram( color="#e9ecef", alpha=0.6, position = 'dodge',bins=20) 
p
```


```{r overlap_prevelance=50}
library(purrr)
list50_decreased <- c(healthysig50, diseasesig_decreased)
paircomp50_decreased <- calcPairwiseOverlaps(list50_decreased)

paircheck50_decreased <- paircomp50_decreased %>%
  filter(stringr::str_detect(name1, 'feces-from-healthy-adult_vs_none---50%-prevalence-threshold')) %>%
  mutate(name2_1=stringr::str_replace(name2, "bsdb:", "")) %>%
  mutate(name2_2=(stringr::str_split(name2_1,"/"))) %>%
  mutate(study_id= purrr::map_chr(name2_2, 1),
          experiment_id= purrr::map_chr(name2_2, 2),
          signature_id= purrr::map_chr(stringr::str_split(purrr::map_chr(name2_2, 3),'_'),1)
        ) 

list50_increased <- c(healthysig50, diseasesig_increased)
paircomp50_increased <- calcPairwiseOverlaps(list50_increased)

paircheck50_increased <- paircomp50_increased %>%
  filter(stringr::str_detect(name1, 'feces-from-healthy-adult_vs_none---50%-prevalence-threshold')) %>%
  mutate(name2_1=stringr::str_replace(name2, "bsdb:", "")) %>%
  mutate(name2_2=(stringr::str_split(name2_1,"/"))) %>%
  mutate(study_id= purrr::map_chr(name2_2, 1),
          experiment_id= purrr::map_chr(name2_2, 2),
          signature_id= purrr::map_chr(stringr::str_split(purrr::map_chr(name2_2, 3),'_'),1)
        ) 
```
## merge jacard scores with disease study data

```{r merge jacard scores with disease study data}
overlap50_decreased<- merge(x = bugall_decreased, y = paircheck50_decreased, by =c('study_id','experiment_id','signature_id'), all.x = TRUE) %>%
  mutate(type="decreased")%>%
  filter(length_set2>4) #keep only bugsigdb studies that tested more than 4 increased or decreased
overlap50_decreased[is.na(overlap50_decreased)] <- 0  

overlap50_increased<- merge(x = bugall_increased, y = paircheck50_increased, by =c('study_id','experiment_id','signature_id'), all.x = TRUE) %>%
  mutate(type="increased") %>%
  filter(length_set2 > 4) #keep only bugsigdb studies that tested more than 4 increased or decreased
overlap50_increased[is.na(overlap50_increased)] <- 0  
```

```{r comparison-increased_vs_decrease_at_prevelance50}
t.test(overlap50_increased$overlap, overlap50_decreased$overlap)
wilcox.test(overlap50_increased$overlap, overlap50_decreased$overlap)

p <- rbind(overlap50_increased,overlap50_decreased) %>%
  ggplot( aes(x=overlap, fill=type)) +
    geom_histogram( color="#e9ecef", alpha=0.6, position = 'dodge',bins=20) 
p
```

```{r overlap_prevelance=70}
library(purrr)
list70_decreased <- c(healthysig70, diseasesig_decreased)
paircomp70_decreased <- calcPairwiseOverlaps(list70_decreased)

paircheck70_decreased <- paircomp70_decreased %>%
  filter(stringr::str_detect(name1, 'feces-from-healthy-adult_vs_none---70%-prevalence-threshold')) %>%
  mutate(name2_1=stringr::str_replace(name2, "bsdb:", "")) %>%
  mutate(name2_2=(stringr::str_split(name2_1,"/"))) %>%
  mutate(study_id= purrr::map_chr(name2_2, 1),
          experiment_id= purrr::map_chr(name2_2, 2),
          signature_id= purrr::map_chr(stringr::str_split(purrr::map_chr(name2_2, 3),'_'),1)
        ) 

list70_increased <- c(healthysig70, diseasesig_increased)
paircomp70_increased <- calcPairwiseOverlaps(list70_increased)

paircheck70_increased <- paircomp70_increased %>%
  filter(stringr::str_detect(name1, 'feces-from-healthy-adult_vs_none---70%-prevalence-threshold')) %>%
  mutate(name2_1=stringr::str_replace(name2, "bsdb:", "")) %>%
  mutate(name2_2=(stringr::str_split(name2_1,"/"))) %>%
  mutate(study_id= purrr::map_chr(name2_2, 1),
          experiment_id= purrr::map_chr(name2_2, 2),
          signature_id= purrr::map_chr(stringr::str_split(purrr::map_chr(name2_2, 3),'_'),1)
        ) 
```
## merge jacard scores with disease study data

```{r merge_jacard_scores_with_disease_study_data}
overlap70_decreased<- merge(x = bugall_decreased, y = paircheck70_decreased, by =c('study_id','experiment_id','signature_id'), all.x = TRUE) %>%
  mutate(type="decreased")%>%
  filter(length_set2>4) #keep only bugsigdb studies that tested more than 4 increased or decreased
overlap70_decreased[is.na(overlap70_decreased)] <- 0  

overlap70_increased<- merge(x = bugall_increased, y = paircheck70_increased, by =c('study_id','experiment_id','signature_id'), all.x = TRUE) %>%
  mutate(type="increased") %>%
  filter(length_set2 > 4) #keep only bugsigdb studies that tested more than 4 increased or decreased
overlap70_increased[is.na(overlap70_increased)] <- 0  
```

```{r comparison_increased_vs_decrease_at_prevelance=70, eval = FALSE}
t.test(overlap70_increased$overlap, overlap70_decreased$overlap)
wilcox.test(overlap70_increased$overlap, overlap70_decreased$overlap)

p <- rbind(overlap70_increased,overlap70_decreased) %>%
  ggplot( aes(x=overlap, fill=type)) +
    geom_histogram( color="#e9ecef", alpha=0.6, position = 'dodge',bins=20) 
p
```

